tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-16 paper.tex
Trim acks

\subsection{Implementation}
\texttt{Tstrait} is written entirely in Python. Many numerical operations are
peformed using standard array-oriented approaches and therefore
highly efficient~\citep{numpy}. Generating the genetic values for a given
trait is computationally intensive. This is done by a
straightforward tree traversal algorithm, which is
accelerated using the \texttt{numba} JIT compiler for Python yielding
excellent performance.

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
For comparison, we also report the CPU time required by a
modified version of the code used by \citet{martin2017} for a
subset of these simulated ARGs.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the VCFs involved are very large, and these tools assume that
the entire dataset fits in RAM.
[DISCUSS RAM USAGE]

[HERE]
The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
VCF, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.

@@

\section{Conclusion}

We believe that in the coming years, simulation in tree sequence encoding will
attract more users in need of conducting genetic simulations with complex
population structure. Multiple studies highlight the advantages of using the
ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
ARG data
structure is becoming increasingly useful in genetic studies, as it is possible
to accurately infer biobank-scale genealogies from sequencing data
\citep{zhang2023}, and simulate large sample whole-genome sequences with
spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
real opportunity for people to further understand the effects of complex
population structure on quantitative traits. The \texttt{tstrait} package is an
essential piece of infrastructure that lets people explore these possibilities.

@@

\section{Acknowledgments}

We greatly acknowledge helpful discussions about quantitative trait simulation
with Gregor Gorjanc. We thank Ben Haller and people in the tskit community for
helpful discussions and feedback. We thank Ben Jeffery for his help with the
Python package development.

>>>>

\subsection{Implementation and validation}
\texttt{Tstrait} is written entirely in Python. Many numerical operations are
peformed using standard array-oriented approaches and therefore
highly efficient~\citep{numpy}. Generating the genetic values for a given
trait is computationally intensive. This is done by a
straightforward tree traversal algorithm, which is
accelerated using the \texttt{numba} JIT compiler for Python yielding
excellent performance.

The \texttt{tstrait} codebase includes an extensive suite of unit tests,
which are automatically run as part of the development process. The
output of \texttt{tstrait} has been statistically validated against
[FILL ME IN].

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
For comparison, we also report the CPU time required by a
modified version of the code used by \citet{martin2017} for a
subset of these simulated ARGs.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the VCFs involved are very large, and these tools assume that
the entire dataset fits in RAM.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest for all these simulations.

@@

\section{Conclusion}
We believe that in the coming years, simulation in tree sequence encoding will
attract more users in need of conducting genetic simulations with complex
population structure. Multiple studies highlight the advantages of using the
ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
ARG data
structure is becoming increasingly useful in genetic studies, as it is possible
to accurately infer biobank-scale genealogies from sequencing data
\citep{zhang2023}, and simulate large sample whole-genome sequences with
spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
real opportunity for people to further understand the effects of complex
population structure on quantitative traits. The \texttt{tstrait} package is an
essential piece of infrastructure that lets people explore these possibilities.

@@

\section{Acknowledgments}
We are grateful to Gregor Gorjanc, Ben Haller, Ben Jeffery and
others in the tskit community for helpful discussions and feedback.
